---
title: "9. Causal Effects and Adjusted Marginal Contrasts (RMST)"
description: "Estimating absolute treatment effects and covariate-adjusted group differences using G-computation."
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{9. Causal Effects and Adjusted Marginal Contrasts (RMST)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Moving Beyond the Hazard Ratio

In clinical trials and observational studies, researchers often wish to compare survival outcomes between two groups. Historically, this is answered using the Hazard Ratio (HR) from a Cox Proportional Hazards model. However, the HR is non-collapsible—meaning the omission of unmeasured covariates will mathematically bias the effect toward the null—and strictly relies on the proportional hazards assumption. If survival curves cross, the HR becomes mathematically invalid.

`SuperSurv` solves this by evaluating group differences on the absolute time scale using the **Restricted Mean Survival Time (RMST)** via G-computation (Standardization) on top of our Ensemble Super Learner. 



RMST calculates the area under the survival curve up to a specific time horizon, $\tau$. By comparing the expected RMST if *everyone* in the dataset belonged to Group 1 versus if *everyone* belonged to Group 0, we obtain a robust, absolute measure of the difference:

$$ \Delta \text{RMST} = E[Y(1)] - E[Y(0)] = \text{RMST}_{\text{Group 1}}(\tau) - \text{RMST}_{\text{Group 0}}(\tau) $$

## Philosophy: "Causal Effect" vs. "Marginal Contrast"

How you interpret this $\Delta \text{RMST}$ depends entirely on the nature of your exposure variable. The math of G-computation is identical for both, but the statistical terminology must be used responsibly.



1. **Causal Average Treatment Effect (ATE):** You can claim a *Causal Effect* if your variable is a **manipulable intervention**. Examples include administering a drug, performing a surgery, or applying a policy. 
   * *Interpretation:* "Administering this drug causally adds an average of 4.2 months of life over a 5-year period compared to the placebo."
2. **Adjusted Marginal Contrast:** You must claim an *Adjusted Marginal Contrast* if your variable is an **immutable trait or biological group**. Examples include biological sex, race, or a genetic biomarker. Because you cannot "causally" intervene to change someone's genetics, we are simply comparing two groups while rigorously adjusting for all other confounding variables.
   * *Interpretation:* "After adjusting for all baseline clinical covariates, the presence of this biomarker is marginally associated with 4.2 additional months of survival over a 5-year period."

## Estimating the Effect with `SuperSurv`

Let's demonstrate this using the built-in `metabric` dataset. We will evaluate the effect of the binary biomarker **`x4`** (1 = present, 0 = absent). Because `x4` is a biomarker, we will interpret the result as an **Adjusted Marginal Contrast**.

```{r setup, message=FALSE, warning=FALSE}
library(SuperSurv)
set.seed(123)

# Load built-in data
data("metabric", package = "SuperSurv")

# Define predictors and time grid
X <- metabric[, grep("^x", names(metabric))]
new.times <- seq(10, 150, by = 10)
```

### 1. Train the Super Learner
First, train the ensemble. We must set `control = list(saveFitLibrary = TRUE)` so the models are saved for the G-computation prediction phase. 

```{r train-model, eval=FALSE}
fit <- SuperSurv(
  time = metabric$duration,
  event = metabric$event,
  X = X,
  newdata = X,
  new.times = new.times,
  event.library = c("surv.coxph", "surv.rfsrc"),
  cens.library = c("surv.coxph"),
  control = list(saveFitLibrary = TRUE) 
)
```


### 2. Estimate the Adjusted Marginal RMST Contrast

We use the `estimate_marginal_rmst()` function to compute an adjusted marginal contrast on the RMST scale. The function sets the binary grouping variable `x4` to 1 for all patients, predicts their survival curves, integrates those predictions up to the restriction horizon `tau`, and then repeats the same procedure with `x4` set to 0. The difference between these two standardized averages yields the adjusted RMST contrast.

```{r causal-effect}
# Estimate the adjusted difference up to tau = 100 months
results <- estimate_marginal_rmst(
  fit = fit, 
  data = metabric, 
  trt_col = "x4", 
  times = new.times, 
  tau = 100
)



print(results$ATE_RMST)
```

**Interpretation:** If the resulting $\Delta$RMST value is `-1.24`, this indicates that, after standardizing over the observed covariate distribution using the fitted Super Learner ensemble, the group with `x4 = 1` is predicted to have approximately 1.24 fewer months of restricted mean survival than the group with `x4 = 0` over a 100-month horizon.


**Uncertainty:** To quantify uncertainty, `estimate_marginal_rmst()` can optionally apply a perturbation-based inference procedure conditional on the fitted ensemble. This returns a perturbation-based standard error, confidence interval, and Wald-type p-value.

```{r}
rmst_results_inf <- estimate_marginal_rmst(
  fit = fit,
  data = metabric,
  trt_col = "x4",
  times = new.times,
  tau = 100,
  inference = TRUE,
  B = 100,
  seed = 123
)

rmst_results_inf$ATE_RMST
rmst_results_inf$SE_RMST
rmst_results_inf$CI_RMST
format.pval(rmst_results_inf$p_value, digits = 3, eps = 1e-16)
```


*Note:* Because this perturbation procedure conditions on the final fitted SuperSurv model and does not refit the learner library or ensemble weights, the resulting confidence interval reflects conditional uncertainty for the standardized RMST contrast and may be relatively narrow.



### 3. Visualizing the Effect Over Time
The difference between groups might be near zero early on but substantial later. We can visualize how the adjusted RMST contrast evolves across different restriction times using `plot_marginal_rmst_curve()`. When `inference = TRUE`, the function also displays perturbation-based confidence intervals as a ribbon.

```{r plot-curve}
# Plot the Delta RMST across a sequence of tau values
tau_grid <- seq(20, 140, by = 30)
plot_marginal_rmst_curve(
  fit = fit, 
  data = metabric, 
  trt_col = "x4", 
  times = new.times, 
  tau_seq = tau_grid,
  inference = TRUE, 
  B = 100, 
  seed = 123, 
  ci_level = 0.95
)
```

### 4. Diagnostic: Predicted RMST vs. Observed Time
To evaluate how well our model's restricted expectations align with reality, we can plot the predicted RMST for the observed data against their true survival times. Patients who experienced the event should lie close to the diagonal line up to $\tau$.



```{r plot-obs}
plot_rmst_vs_obs(
  fit = fit, 
  data = metabric, 
  time_col = "duration", 
  event_col = "event", 
  times = new.times, 
  tau = 350
)
```
